Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-22 second-draft/pnas-word-order-si.tex
SI

Here we discuss how our formalization of communicative efficiency relates to other information-theoretic work on language.
Our formalization of efficiency is based on the function proposed in \cite{ferreri2003least,ferrericancho2007global,futrell2017memory,peloquin2019interactions} as a general efficiency metric for communicative systems.
The key idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms. %; there are some differences in the formalization of complexity.
Formally, if $S$ denotes signals (e.g., words, sentences) and $R$ denotes their referents (e.g., objects in a reference game), then this efficiency metric takes the form (notation slightly varies across these publications):
\begin{equation}\label{eq:eff-general}
I[S, R] - \lambda H[S]
\end{equation}
where $I[S, R]$ describes the \key{informativity} of the signals $S$ about their referents $R$, and $H[S]$ describes the \key{complexity} of the communication system.
While prior studies \cite{ferreri2003least, kemp2012kinship,regier2015word,zaslavsky2018efficient} mostly considered settings where the signals $S$ are individual words without further structure, the signals are entire sentences $\mathcal{U}$ in our setting.
The underlying messages $R$ which the speaker aims to convey are the syntactic structures $\mathcal{T}$.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The syntactic structure specifies how the meanings of words are combined; therefore, recovering the syntactic structure is a prerequisite to understanding a sentence correctly.

@@

@@

Figure (REF) in the main paper shows the average of these per-language plots. \textcolor{red}{unify pareto curve computation in main paper}

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} % figures/pred-eff-families-2.pdf
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language. \textcolor{red}{TODO fix order of 1-8}}\label{fig:per-lang}
%	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by efficiency optimization (right). Dark blue: Verb patterner \emph{precedes} object patterner(English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...) \textcolor{red}{add other functions to this plot}.}\label{fig:per-lang}
\end{figure}

@@

\paragraph{Bayesian Regression}
We modeled the

We modeled the probabilities $p_{i,j}$ that a grammar $g_i$ satisfies the $j$-th correlation ($j=1,...,8$) using a multilevel logistic model \textcolor{red}{TODO (CITE)}, with random intercepts for the language for whose data the grammar had been optimized, and its language family, annotated according to \url{http://universaldependencies.org/}..
\textcolor{red}{TODO some more details here}
%See SI Section S4.3 for details.

As we constructed 4 verb-object and 4 object-verb grammars for each of the 51 languages, TODO...

Following the recommendations of \cite{ghosh2018use, burkner2018advanced}, we used as a very weakly informative prior a Student's $t$ prior with $\nu=3$ degrees of freedom, mean 0, and scale $\sigma=10$ (i.e., the PDF $p$ is $\frac{1}{\sigma} p_3(x/\sigma)$, where $p_3$ is the PDF of the $t$-distribution with $\nu=3$).
Note that a correlation that holds in 90\% of cases corresponds to an intercept $\alpha \approx 2.19$ in the logistic model, well within the main probability mass of the prior.
The covariance matrix is modeled using the LKJ prior, as described in \cite{burkner2018advanced}.
We used four chains, with 5000 samples each, of which the first 2500 were discarded as warmup samples.
We confirmed convergence using $\hat{R}$ and visual inspection of chains (REF).

@@

\paragraph{Density Plots}
We obtained the posterior density plots in Figures (REF main, REF SI) by applying the logistic transformation ($x \mapsto \frac{1}{1+\exp(-x)}$) to the posterior samples of $\beta$.

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu} we report the results of the simulations we did before the main experiment reported in the main paper (left and center), together with their preregistrations, besides the results from the main experiment (right).
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) the first simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.
\textcolor{red}{TODO also the aggregate posterior}

@@

@@

Note that the the derivatives inside the expectations on the right hand sides can all be computed using backpropagation for our neural network architectures (CITE).

@@

\paragraph{Estimating Predictability}
We choose a standard LSTM \citep{hochreiter1997long} language model with vocabulary size restricted to the most frequent 50,000 words in the treebanks for a given language.
Given the small size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its POS tags in each step.
Using POS tags is intended to prevent overfitting on small corpora.

@@

To reduce overfitting on small corpora, we choose a delexicalized setup, parsing only from POS tags. Preliminary experiments showed that a parser incorporating word forms overfitted long before the ordering model had converged; parsing from POS tags prevents early overfitting.
This decision was made before evaluating word order properties.

@@

We estimated probabilities by taking counts in the training partition, and performing Laplace smoothing with a pseudocount $\alpha=1$ for each possible rule of this form.
For such a PCFG, exact parsing is possible using Dynamic Programming, and specifically the CKY algorithm \cite{kasami1966efficiency}.

>>>>

Here we discuss how our formalization of communicative efficiency relates to other information-theoretic work on language.
Our formalization of efficiency is based on the function proposed in \cite{ferreri2003least,ferrericancho2007global,futrell2017memory,peloquin2019interactions} as a general efficiency metric for communicative systems.
The key idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms. %; there are some differences in the formalization of complexity.
Formally, if $S$ denotes signals (e.g., words, sentences) and $R$ denotes their referents (e.g., objects in a reference game), then this efficiency metric takes the form (notation slightly varies across these publications):
\begin{equation}\label{eq:eff-general}
I[S, R] - \lambda H[S]
\end{equation}
where $I[S, R]$ describes the \key{informativity} of the signals $S$ about their referents $R$, and $H[S]$ describes the \key{complexity} of the communication system.
While prior studies \cite{ferreri2003least, kemp2012kinship,regier2015word,zaslavsky2018efficient} mostly considered settings where the signals $S$ are individual words without further structure, the signals are entire sentences $\mathcal{U}$ in our setting.
The underlying messages $R$ which the speaker aims to convey are the syntactic structures $\mathcal{T}$.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The syntactic structure specifies how the meanings of words are combined; therefore, recovering the syntactic structure is a prerequisite to understanding a sentence correctly.
%The idea that efficient languages will permit the listener to infer grammati

@@

%\textcolor{red}{also mention \cite{piantadosi2012communicative,fedzechkina2012language}}

@@

Figure (REF) in the main paper shows the average of these per-language plots.

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} % figures/pred-eff-families-2.pdf
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language.}\label{fig:per-lang}
%	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by efficiency optimization (right). Dark blue: Verb patterner \emph{precedes} object patterner(English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...) .}\label{fig:per-lang}
\end{figure}

@@

\paragraph{Bayesian Regression}
We modeled the probabilities $p_{L,j}$ that a grammar optimized for data from language $L$ satisfies the $j$-th correlation ($j=1,...,8$) using a multilevel logistic model \cite{gelman2013bayesian}, with random intercepts for the language for whose data the grammar had been optimized, and for its language family, annotated according to \url{http://universaldependencies.org/}.
Formally,
\begin{equation}\label{eq:mixed-effects}
logit(p_{L,j}) = \alpha_j + u_{L,j} + v_{f_L,j}
\end{equation}
where $f_L$ is the language family of $L$.
The intercepts $\alpha_j$ ($j=1,...8$) encode the population-level prevalence of the correlations when controlling for differences between datasets from different languages and language families; $u_{L,j}$, $v_{f_L,j}$ encode per-language and per-family deviations from the population-level intercept $\alpha_j$.
%, and
%$u_{L,1,...8} \sim \mathcal{N}(0, \Sigma_{lang})$,
%$v_{f_L,1,...8} \sim \mathcal{N}(0, \Sigma_{fam})$.

%As we constructed 4 verb-object and 4 object-verb grammars for each of the 51 languages, TODO...

Following the recommendations of \cite{ghosh2018use, burkner2018advanced}, we used as a very weakly informative prior a Student's $t$ prior with $\nu=3$ degrees of freedom, mean 0, and scale $\sigma=10$ (i.e., the PDF $p$ is $\frac{1}{\sigma} p_3(x/\sigma)$, where $p_3$ is the PDF of the $t$-distribution with $\nu=3$).
We used this prior for $\alpha_j, \sigma_{L,j}, \tau_{L,j}$.
Note that a correlation that holds in 90\% of cases corresponds to an intercept $\alpha \approx 2.19$ in the logistic model, well within the main probability mass of the prior.

We modeled full covariance matrices of per-language and per-family random intercepts over all eight correlations. We placed an LKJ prior ($\eta=1$) on these matrices, as described in \cite{burkner2018advanced}.
We used MCMC sampling implemented in Stan \cite{carpenter2017stan, hoffman2014no} using the R package \texttt{brms} \cite{buerkner2017brms}.
We ran four chains, with 5000 samples each, of which the first 2500 were discarded as warmup samples.
We confirmed convergence using $\hat{R}$ and visual inspection of chains \cite{gelman2013bayesian}.

We obtained the posterior density plots in Figures (REF main, \ref{tab:all-predictions-1}) by applying the logistic transformation ($x \mapsto \frac{1}{1+\exp(-x)}$) to the posterior samples of $\alpha_j$ (\ref{eq:mixed-effects}).

@@

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu} we report the results of the simulations we did before the main experiment reported in the main paper (left and center), together with their preregistrations, besides the results from the main experiment (right).
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) the first simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.

\textcolor{red}{TODO decide on formatting}

@@

\begin{figure}[ht]
	\begin{center}
	\includegraphics[width=0.25\textwidth]{../results/correlations/figures/posterior-satisfied-universals-together-large-prior-efficiency09.pdf}
	\includegraphics[width=0.25\textwidth]{../results/correlations/figures/posterior-satisfied-universals-together-large-prior-efficiency10.pdf}
	\end{center}
	\caption{(Other experiments) Posterior of the number of correlations correctly predicted by efficiency and its components, in a Bayesian multivariate mixed-effects logistic regression with random effects for languages and language families.}\label{fig:posterior-prior}
\end{figure}

@@

Note that the the derivatives inside the expectations on the right hand sides can all be computed using backpropagation for our neural network architectures.

@@

\paragraph{Estimating Predictability}
We choose a standard LSTM language model \citep{goldberg2017neural, hochreiter1997long}.
We restrict the vocabulary to the most frequent 50,000 words in the treebanks for a given language.
Given the moderate size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its POS tags in each step.
Using POS tags is intended to prevent overfitting on small corpora.

@@

To reduce overfitting on small corpora, we choose a delexicalized setup, parsing only from POS tags. Preliminary experiments showed that a parser incorporating word forms overfitted long before the ordering grammar had converged; parsing from POS tags prevents early overfitting.
This decision was made before evaluating word order properties.

@@

We estimated probabilities by taking counts in the training partition, and performing Laplace smoothing with a pseudocount $\alpha=1$ for each possible rule of this form.
For such a PCFG, exact parsing is possible using Dynamic Programming, and specifically the CKY algorithm \cite{kasami1966efficient}.
